Detecting somatic variants at sub-1\% allele frequency in liquid biopsy samples
requires accurate UMI-based error suppression and sensitive variant calling.
Current pipelines chain several tools that discard molecule-level information
between steps, making per-molecule strand bias and duplex evidence unavailable
at the calling stage.

We present kam, an alignment-free variant detection pipeline for Twist duplex
UMI panel sequencing. kam assembles duplex molecules directly from raw FASTQ,
indexes k-mers with per-molecule provenance, builds a de Bruijn graph over all
observed k-mers, and calls variants by scoring alternative paths against the
reference path using molecule counts and strand composition. No reference
alignment is performed. Called variants are left-normalised to standard VCF
convention, enabling direct comparison against truth sets for SNVs, indels, and
structural variants. Structural variant detection uses junction k-mer injection:
breakpoint-spanning k-mers are added to the allowlist, making SV allele paths
discoverable in the de Bruijn graph without specialised alignment.

We evaluate kam in two operating modes on the Twist cfDNA Pan-Cancer Reference
Standard v2 titration series (375 truth variants, 24 samples, 5--30\,ng input
DNA, VAF 0\%--2\%). In discovery mode, kam detects 54--63\% of truth variants
at 2\% VAF with precision 0.69--0.84 using 2M read pairs per sample. In
tumour-informed monitoring mode (\texttt{--target-variants}), the same calls
are filtered to variants matching a pre-specified target VCF: precision rises
to 1.0 at all VAF levels with no change to sensitivity.

We identified and fixed two implementation bugs during evaluation: a strand
bias overcounting bug that inflated the Fisher exact test sample size
approximately 70-fold, and a duplex orientation comparison bug that prevented
genuine duplex pairs from being grouped correctly.
With both fixes applied and 2M read pairs per sample, duplex identification
rates are 6.2--18.9\% of molecules. With recommended filters
(\texttt{--max-vaf 0.35 --min-family-size 2}) and tumour-informed mode applied,
kam detects 54--63\% of truth variants at 2\% VAF with precision 1.0, running
end-to-end in 41--55 seconds per sample on a single core. SNV sensitivity
reaches 69--80\% at 2\% VAF; indel sensitivity is 32--39\%. Sensitivity at
0.25--0.5\% VAF reaches 26--46\% at 15--30\,ng input and 6--17\% at 5\,ng input.

kam is implemented in Rust and is available as an open-source workspace.
