\section{Discussion}
\label{sec:discussion}

kam is a proof of concept that alignment-free variant detection is feasible for
Twist duplex UMI panel sequencing.
The pipeline preserves molecule-level information from raw FASTQ to variant call
without any alignment step, detecting 52--61\% of truth variants at 2\% VAF
with precision 0.58--0.78 using 2M read pairs per sample, running in 19--35
seconds on a single core.

Two implementation bugs were identified and fixed during evaluation: a strand
bias overcounting bug that inflated the Fisher exact test sample size
approximately 70-fold, and a duplex orientation comparison bug that prevented
genuine duplex pairs from being grouped correctly.
With both fixes applied, sensitivity extends to the 0.25--0.5\% VAF range at
15--30\,ng input.
A germline VAF filter (\texttt{--max-vaf 0.35}) reduces false positive counts
by 8--11 per sample by removing heterozygous germline SNPs at VAF $\approx 0.5$;
this filter is recommended for somatic variant calling in cfDNA where matched
normal is not available.
The remaining false positive floor (50--96 per sample) reflects background
biological signal in the cfDNA that is indistinguishable from spike-in calls
without external annotation.
The primary depth limitation is the O($n^2$) UMI grouping step, which caps
practical depth at approximately 2M read pairs within a 4\,GB memory budget.

\paragraph{SNV and indel calling.}
Indel sensitivity at 2M reads ranges from 32\% to 39\% at 2\% VAF, compared to
69--80\% for SNVs. The gap is expected: insertions and deletions shift the
position of end-anchor k-mers relative to the reference, requiring those
shifted k-mers to be covered by sequencing reads for a walk to complete. This
anchor coverage requirement is more likely to be satisfied as depth increases.
The left-normalisation step ensures that called indel positions match the
convention used by GATK, bcftools, and standard truth sets~\autocite{tan2015},
enabling direct comparison against VCF truth files.

\paragraph{Depth scaling.}
Sensitivity grows with read depth across the full 250K--2M range tested.
Gains are largest from 250K to 1M reads; returns diminish above 1M at the
high-VAF end. This is consistent with a model where the most abundant variant
molecules are covered first, and rarer variants require proportionally more
reads. Sub-0.1\% VAF detection requires deeper sequencing; this is feasible
once the O($n^2$) UMI grouping bottleneck is resolved by hash partitioning
(Section~\ref{sec:future}).

\paragraph{Comparison to alignment-based approaches.}
Alignment-based pipelines for ctDNA variant calling typically report sensitivity
of 80--95\% at 1--2\% VAF with comparable panel sizes, using millions of
read pairs~\autocite{Twist22cfdna}. The gap between those numbers and the
52--61\% reported here has two primary components: the anchor coverage
limitation for indels (SNV sensitivity at 2\% VAF reaches 69--80\%, narrowing
the gap considerably), and the conservative minimum alt molecule threshold.
A direct head-to-head comparison on the same samples against GATK Mutect2,
Sentieon TNscope, or fgbio\,+\,bwa is not included in this work; such a
comparison requires running alignment-based callers on matched read sets, which
is deferred to future evaluation (Section~\ref{sec:future}).

\paragraph{Information preservation.}
The key architectural advantage of kam over the km/Jellyfish pipeline is that
every k-mer in the graph retains a pointer to the molecule that contributed it.
This means strand bias, duplex support, and per-molecule error probability are
available at the scoring stage without a separate alignment pass. Current calling
uses only molecule counts and strand counts; the full posterior over molecule
quality (incorporating per-cycle error probability from base quality scores)
is implemented but not yet tuned.

\paragraph{Reproducibility.}
All RNG in the pipeline uses seeded deterministic generators, ensuring that
runs on the same input produce identical output. This is a requirement for
Nextflow cache compatibility and for scientific reproducibility. The pipeline
does not use the system clock, process ID, or any other non-deterministic
seed source.
