\section{Future Work}
\label{sec:future}

\paragraph{Hash-partition UMI grouping for deeper sequencing.}
The UMI read-grouping step has two complexity terms.
The O($n \times u$) linear-scan lookup of read pairs against unique UMI entries has
been replaced with a hash-map giving O($n$) total.
The remaining Hamming-distance clustering step scales as O($u^2$) where $u$ is
the number of unique canonical UMI pairs.
At 2\,M reads this term does not dominate; at 10\,M reads, $u$ approaches the
theoretical maximum of 524\,K pairs for 5-base duplex UMIs and clustering would
become a hard wall.
Partitioning UMIs by their first 3-base prefix (64 buckets) before clustering
reduces this term by 64$\times$ to O($u^2/64$).
These two changes together enable processing of 10--20\,M read pairs, giving
5--10$\times$ more molecules per target and substantially reducing coverage-gap
walk failures at low VAF.

\paragraph{Relaxed end-anchor matching for indels.}
The DFS walk requires an exact k-mer match at the reference end anchor. For an
indel of length $\ell$, the variant path is $\ell$ k-mers shorter or longer than
the reference path, and reads must span from the variant junction to the shifted
anchor position. An investigation of missed variants confirms that end-anchor
coverage is the primary cause of the 61--68\% indel false-negative rate.
Relaxing the end-anchor requirement to allow termination within a $\pm \ell$-kmer
window would recover the majority of missed indels without affecting SNV calls.

Beyond algorithm fixes, the following extensions are planned.

\paragraph{De novo variant discovery.}
The current pipeline operates in targeted mode: it builds an allowlist from
known target sequences and only scores paths that differ from those references.
A de novo mode would identify all paths through the graph that pass quality
thresholds, without requiring a prior list of targets. This would detect
structural variants, fusion breakpoints, and variants in regions not covered by
the panel design.

\paragraph{Nextflow integration.}
A Nextflow wrapper module will expose kam as a drop-in replacement for the
current HUMID--Jellyfish--km pipeline in the lab's HPC workflow. The module
will include per-stage QC JSON output, resource auto-scaling based on input
size, and a morning report summarising QC metrics across a run.

\paragraph{Head-to-head comparison.}
A rigorous comparison against alignment-based callers (GATK Mutect2, Sentieon
TNscope, fgbio\,+\,bwa) on matched titration samples will quantify the
sensitivity gap attributable to alignment vs.\ the gap attributable to depth.
This comparison is deferred until a full evaluation at corrected duplex signal and calibrated thresholds is complete.
